---
title: "Survival Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = F}
library(survival)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(kableExtra)
library(jsonlite)

consts <- read_json(file.path(repo_dir, 'config.json'))
proj_dir <- file.path(consts[['official_indir']], 'TARGET')
figure_outdir <- file.path(repo_dir, 'figures_revision')

clinical <- read.table(file.path(proj_dir, 'clinical.processed.withAges.tsv'), sep='\t', header=T, na.strings = c("", "NA"))
clinical$in_analysis_dataset <- as.logical(clinical$in_analysis_dataset)
clinical$in_analysis_dataset[is.na(clinical$in_analysis_dataset)] <- FALSE
clinical$myc <- clinical$MYCN.status

cat(sprintf('Number of tumors: %d', sum(clinical$in_analysis_dataset)))
```

# Prepare data

```{r kaplan-meier, message = F}
sf = 0.6
label_fontsize <- 23
p_val_size <- 6
  
# Restrict columns
survival_data <- data.frame(
  Patient=clinical$submitter_id,
  event=as.numeric(clinical$vital_status == 'Dead'),
  # time=clinical$days_to_follow_up / 365,
  time=clinical$days_to_last_follow_up / 365,
  mitotic.age=scale(clinical$phi),                # Use phi as tumor mitotic age - scaled
  mitotic.age.split1=scale(clinical$phi.split1),
  mitotic.age.split2=scale(clinical$phi.split2),
  # DELETE
  LUMP=scale(clinical$LUMP),
  #
  inss_stage=clinical$inss_stage,
  cog_neuroblastoma_risk_group=clinical$cog_neuroblastoma_risk_group,
  patient_age=clinical$age_at_diagnosis / 365,
  myc=clinical$myc,
  in_analysis_dataset=clinical$in_analysis_dataset
  )
survival_data <- survival_data[survival_data$in_analysis_dataset, ]

sprintf('%d samples available', nrow(survival_data))

# KM curves

#########################################################
############### stratify by mitotic.age #################
#########################################################

# Halves
survival_data <- survival_data %>%
    mutate(
      dichom = ntile(mitotic.age, 2),
      dichom = factor(dichom, levels = c(1, 2), labels = c("Young", "Old")),
      #
      dichom.split1 = ntile(mitotic.age.split1, 2),
      dichom.split1 = factor(dichom.split1, levels = c(1, 2), labels = c("Young", "Old")),
      #
      dichom.split2 = ntile(mitotic.age.split2, 2),
      dichom.split2 = factor(dichom.split2, levels = c(1, 2), labels = c("Young", "Old")),
      )
```


# Kaplan-meier curves {.tabset .tabset-fade}

## All tumors - no stratification
```{r all-tumors-km-no-strat}
#######################################
########### All tumors ################
#######################################

fit <- survfit(Surv(time, event) ~ 1, data=survival_data)

# Plot with customizations and p-value
km_plot <- autoplot(fit) +
  ylim(0, 1) +
  labs(
    x = "Time since diagnosis (years)",
    y = "Survival",
    # title = sprintf('All tumors (N = %d)', sum(fit$n))
  ) +
  theme_classic() +
  theme(plot.title = element_text(size=sf * label_fontsize, hjust = 0.5),
        axis.title = element_text(size=sf * label_fontsize)
        )

ggsave(file.path(figure_outdir, "km_plot_no_stratification.svg"), km_plot, width = sf * 8, height = sf * 6, dpi = 800, device='svg', units='in')

cat('\nEstimate OS for different time points (in years):\n\n')
summary(fit, times = c(5, 10))

km_plot
```

```{r km_stratified_plot_function}
#######################################
########### All tumors ################
#######################################

makeKmStratifiedPlot <- function(survival_data, group_var, filepath) {
  
  form <- as.formula(paste("Surv(time, event) ~", group_var))
  fit <- survfit(form, data=survival_data)
  
  # Compute p-value
  logrank <- survdiff(form, data = survival_data)
  p_val <- 1 - pchisq(logrank$chisq, df = length(logrank$n) - 1)
  p_label <- paste0("P = ", signif(p_val, 2))
  
  # Plot with customizations and p-value
  km_plot <- autoplot(fit) +
  ylim(0, 1) +
  labs(
    x = "Time since diagnosis (years)",
    y = "Survival",
    # title = sprintf('All tumors (N = %d)', sum(logrank$n))
  ) +
  annotate("text", x = 12, y = 0.9, label = p_label, size = sf * p_val_size) +
  theme_classic() +
  theme(plot.title = element_text(size=sf * label_fontsize, hjust = 0.5),
        axis.title = element_text(size=sf * label_fontsize),
        legend.title = element_blank(),
        legend.text = element_text(size=sf * label_fontsize * 0.8),
        legend.position = c(0.01, 0.03),  # bottom-left corner
        legend.justification = c(0, 0)
        ) +
  scale_color_discrete(labels = c(
    sprintf('Young tumors (n = %d)', logrank$n[[sprintf('%s=Young', group_var)]]),
    sprintf('Old tumors (n = %d)', logrank$n[[sprintf('%s=Old', group_var)]])
  )) +
  scale_fill_discrete(labels = c(
    sprintf('Young tumors (n = %d)', logrank$n[[sprintf('%s=Young', group_var)]]),
    sprintf('Old tumors (n = %d)', logrank$n[[sprintf('%s=Old', group_var)]])
  ))
  
  ggsave(filepath, km_plot, width = sf * 8, height = sf * 6, dpi = 800, device='svg', units='in')

  km_plot
}
```


## All tumors 
```{r all-tumors-km}
#######################################
########### All tumors ################
#######################################

makeKmStratifiedPlot(survival_data, "dichom", file.path(figure_outdir, "km_plot.svg"))
```

## All tumors, sensitivity split1 
```{r all-tumors-split1}
#######################################
########### All tumors ################
#######################################

makeKmStratifiedPlot(survival_data, "dichom.split1", file.path(figure_outdir, "km_plot.split1.svg"))
```

## All tumors, sensitivity split2
```{r all-tumors-split2}
#######################################
########### All tumors ################
#######################################

makeKmStratifiedPlot(survival_data, "dichom.split2", file.path(figure_outdir, "km_plot.split2.svg"))
```

# Cox-models {.tabset .tabset-fade}

```{r helper-func, message=F}
coxResultsTable <- function(mod) {
  modsum <- summary(mod)
  coefs <- as.data.frame(modsum$coefficients) # Get coefficients
  conf.int <- as.data.frame(modsum$conf.int)
  # colnames(coef_table) <- c("Coef", "Exp(Coef)", "SE(Coef)", "z", "p")
  coef_table <- cbind(coefs, conf.int)
  names(coef_table)[which(names(coef_table) == 'Pr(>|z|)')] = 'p'
  # Reformat
  coef_table <- coef_table[, -c(6, 7)]
  # format(coef_table, digits=2)
  coef_table[] <- lapply(coef_table, function(x) {
    if (is.numeric(x)) signif(x, 2) else x
  })
  coef_table
}
```

## Predict only with tumor mitotic age

```{r cox-models, message = F}
surv.model <- coxph(Surv(time, event) ~ mitotic.age,
                                   data = survival_data)
cat(sprintf('Used %d samples\n', summary(surv.model)$n))

coef_table <- coxResultsTable(surv.model)

# Display as a kable table
kable(coef_table, caption = "Cox Proportional Hazards Model Summary") %>%
  kable_styling(full_width = FALSE, position = "left")

schoenfeld_test <- cox.zph(surv.model)
print(schoenfeld_test)

# Plot residuals for proportional hazards check
plot(schoenfeld_test)
```

## Predict only with tumor mitotic age and LUMP

```{r cox-models-2, message = F}
surv.model <- coxph(Surv(time, event) ~ mitotic.age + LUMP,
                                   data = survival_data)
cat(sprintf('Used %d samples\n', summary(surv.model)$n))

coef_table <- coxResultsTable(surv.model)

# Display as a kable table
kable(coef_table, caption = "Cox Proportional Hazards Model Summary") %>%
  kable_styling(full_width = FALSE, position = "left")

schoenfeld_test <- cox.zph(surv.model)
print(schoenfeld_test)

# Plot residuals for proportional hazards check
plot(schoenfeld_test)
```

## Predict with tumor mitotic age, patient age, stage, and MYC amplication

```{r check-confounding-2}
survival_data$stage4 <- survival_data$inss_stage == 'Stage 4'
surv.model <- coxph(Surv(time, event) ~ mitotic.age + stage4 + patient_age + myc, data = survival_data[!is.na(survival_data$myc),])

cat(sprintf('Used %d samples\n', summary(surv.model)$n))

coef_table <- coxResultsTable(surv.model)

# Display as a kable table
kable(coef_table, caption = "Cox Proportional Hazards Model Summary") %>%
  kable_styling(full_width = FALSE, position = "left")


schoenfeld_test <- cox.zph(surv.model)
print(schoenfeld_test)
# Display results

# Plot residuals for proportional hazards check
plot(schoenfeld_test)
```


## Predict with tumor mitotic age, patient age, stage MYC amplication, and LUMP

```{r check-confounding-2-lump}
survival_data$stage4 <- survival_data$inss_stage == 'Stage 4'
surv.model <- coxph(Surv(time, event) ~ mitotic.age + stage4 + patient_age + myc + LUMP, data = survival_data[!is.na(survival_data$myc),])

cat(sprintf('Used %d samples\n', summary(surv.model)$n))

coef_table <- coxResultsTable(surv.model)

# Display as a kable table
kable(coef_table, caption = "Cox Proportional Hazards Model Summary") %>%
  kable_styling(full_width = FALSE, position = "left")


schoenfeld_test <- cox.zph(surv.model)
print(schoenfeld_test)
# Display results

# Plot residuals for proportional hazards check
plot(schoenfeld_test)
```

## Predict only with tumor mitotic age - sensitivity split1

```{r cox-models-split1, message = F}
surv.model <- coxph(Surv(time, event) ~ mitotic.age.split1,
                                   data = survival_data)
cat(sprintf('Used %d samples\n', summary(surv.model)$n))

coef_table <- coxResultsTable(surv.model)

# Display as a kable table
kable(coef_table, caption = "Cox Proportional Hazards Model Summary") %>%
  kable_styling(full_width = FALSE, position = "left")

schoenfeld_test <- cox.zph(surv.model)
print(schoenfeld_test)

# Plot residuals for proportional hazards check
plot(schoenfeld_test)
```

## Predict with tumor mitotic age, patient age, stage, and MYC amplication - sensitivity split1

```{r check-confounding-2-split1}
survival_data$stage4 <- survival_data$inss_stage == 'Stage 4'
surv.model <- coxph(Surv(time, event) ~ mitotic.age.split1 + stage4 + patient_age + myc, data = survival_data[!is.na(survival_data$myc),])

cat(sprintf('Used %d samples\n', summary(surv.model)$n))

coef_table <- coxResultsTable(surv.model)

# Display as a kable table
kable(coef_table, caption = "Cox Proportional Hazards Model Summary") %>%
  kable_styling(full_width = FALSE, position = "left")


schoenfeld_test <- cox.zph(surv.model)
print(schoenfeld_test)
# Display results

# Plot residuals for proportional hazards check
plot(schoenfeld_test)
```

## Predict only with tumor mitotic age - sensitivity split2

```{r cox-models-split2, message = F}
surv.model <- coxph(Surv(time, event) ~ mitotic.age.split2,
                                   data = survival_data)
cat(sprintf('Used %d samples\n', summary(surv.model)$n))

coef_table <- coxResultsTable(surv.model)

# Display as a kable table
kable(coef_table, caption = "Cox Proportional Hazards Model Summary") %>%
  kable_styling(full_width = FALSE, position = "left")

schoenfeld_test <- cox.zph(surv.model)
print(schoenfeld_test)

# Plot residuals for proportional hazards check
plot(schoenfeld_test)
```

## Predict with tumor mitotic age, patient age, stage, and MYC amplication - sensitivity split2

```{r check-confounding-2-split2}
survival_data$stage4 <- survival_data$inss_stage == 'Stage 4'
surv.model <- coxph(Surv(time, event) ~ mitotic.age.split2 + stage4 + patient_age + myc, data = survival_data[!is.na(survival_data$myc),])

cat(sprintf('Used %d samples\n', summary(surv.model)$n))

coef_table <- coxResultsTable(surv.model)

# Display as a kable table
kable(coef_table, caption = "Cox Proportional Hazards Model Summary") %>%
  kable_styling(full_width = FALSE, position = "left")


schoenfeld_test <- cox.zph(surv.model)
print(schoenfeld_test)
# Display results

# Plot residuals for proportional hazards check
plot(schoenfeld_test)
```
