---
title: "Purity Adjustment and CIBERSORT"
author: "Shannon T. Holloway"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: "hide"
params:
  save_path: "TARGET"
  methyl_filename: "cohort1.methyl"
  gene_filename: "cohort1.rnaseq"
  processed_clinical_filename: "clinical.processed.tsv"
  lookup_filename: "xlsx file containing CpG site specific beta values from WB"
  cibersort_filename: "Cibersort_LM22_500perm_Analysis_Tumors.csv"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
library(ggplot2)
library(ggpubr)
library(ggsci)
library(jsonlite)

source('util.R')

consts <- read_json(file.path(repo_dir, 'config.json'))

if (consts[["Windows"]]) {
  save_path <- file.path("P:", params$save_path)
} else {
  save_path <- file.path(consts[['official_indir']], params$save_path)
}

gg_theme <- theme(panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank(),
                  panel.background = element_blank(), 
                  axis.line = element_line(colour = "black"))
```

#  {.tabset}

## Data {.tabset}

### Read TCGA Methylation Data

```{r read_methylation_from_gdc}
brca.methyl <- tryCatch(readRDS(file.path(save_path, 
                                          paste0(params$methyl_filename, ".rds"))),
                        error = function(e) {
                          stop("must recreate the tumor methylation object")
                          })
cat("Dimensions of raw methylation data", dim(brca.methyl), "\n")
```

### Restrict to Analysis Tumors

We keep only the tumors in the analysis cohort that have bulk RNA data, which
are the tumors included in the CIBERSORT analysis.

```{r restrict_to_analysis_tumors}
clinical <- read.table(file.path(save_path, 
                                 params$processed_clinical_filename), 
                       sep = "\t", header = TRUE)
clinical$in_analysis_dataset <- as.logical(clinical$in_analysis_dataset)
analysis_tumors <- clinical$sampleID[clinical$in_analysis_dataset]

cibersort <- read.csv(file.path(save_path, 
                                params$cibersort_filename))

methyl_tumors <- getTumorIDs(colnames(brca.methyl))
keep <- methyl_tumors %in% analysis_tumors & 
  methyl_tumors %in% getTumorIDs(cibersort$Mixture)
brca.methyl <- brca.methyl[, keep]

cat("Tumors in the analysis cohort", length(analysis_tumors), "\n")
cat("Dimensions of current methylation data", dim(brca.methyl), "\n")
```


### Read Bulk RNA Data

```{r read_rna_from_gdc}
brca.seq <- readRDS(file.path(save_path, 
                              paste0(params$gene_filename, "_raw_counts.rds")))
cat("Dimensions of raw RNAseq data", dim(brca.seq), "\n")

brca.seq_ensembl <- brca.seq$gene_id
brca.seq_gene.name <- brca.seq$gene_name

idx <- match(getTumorIDs(colnames(brca.methyl)),
             getTumorIDs(colnames(brca.seq)))
if (any(is.na(idx))) stop("something went wrong in RNA data!")
brca.seq <- brca.seq[, idx]
brca.seq <- data.matrix(brca.seq)

cat("Dimensions of current RNAseq data", dim(brca.seq), "\n")
```

#### Read CIBERSORTx Results

The CIBERSORT analysis was completed for the tumors in the analysis cohort with
bulk RNA data using the LM22 signature, B-mode batch correction, absolute
mode, quantile normalization disabled, and 500 permutations.

```{r cibersor_read}
idx <- match(getTumorIDs(colnames(brca.methyl)), getTumorIDs(cibersort$Mixture))
cibersort <- cibersort[idx[!is.na(idx)], ]
cat("All current tumors are contained in the CIBERSORT results", 
    isTRUE(all.equal(getTumorIDs(cibersort$Mixture), getTumorIDs(colnames(brca.methyl)))),
    "\n")
cat("Dimensions of CIBERSORT results", dim(cibersort), "\n")
```

### Purity Values

Purity estimates are the LUMP values in the processed clinical table. The
LUMP method is taken from Aran, D., Sirota, M. and Butte, A. J. (2015)
Systematic pan-cancer analysis of tumor purity. Nature Communications,
6, 8971.

```{r read_purity}
# keep only the data for tumors used in the analysis
idx <- match(getTumorIDs(colnames(brca.methyl)), clinical$sampleID)
cat("All current tumors are in the clinical table", !any(is.na(idx)), "\n")
purity <- data.frame("Sample.ID" = clinical$sampleID[idx],
                     "LUMP" = clinical$LUMP[idx])
```

### Read CpG Sites

```{r clock_set_data}
cpg_sites <- read.csv(file.path(repo_dir, "3. Select fCpGs", "outputs", 
                                "NBL_Clock_CpGs.txt"),
                      header = FALSE)[, 1L]
```

We subset the methylation data to only the CpG sites of interest

```{r limit_cpg_site}
brca.methyl <- brca.methyl[rownames(brca.methyl) %in% cpg_sites, ]
cat("Dimensions of current methylation data", dim(brca.methyl), "\n")
```

## Methylation Parameters {.tabset}

### Intro

The methylation parameters we consider here are calculated from:

-   the measured $\beta$ ($c_{\beta}$),
-   the "model 1" purity adjusted $\beta$ ($c_{{\widehat{\beta}}}^{\alpha}$)

### Measured ($c_{\beta}$) {.tabset}

#### Definition

The methylation parameter for tumor $i$, $s_{\beta, i}$, is the standard
deviation of the $\beta$-values of the selected CpG sites.

$$
s^2_{\beta,i} = \frac{1}{(n_{CPG}-1)} \sum_{j = 1}^{500} (\beta_{j,i} - \mu_{\beta,i})^2 \mathrm{I}[!\mathrm{is.na}(\beta_{j,i})],
$$ where
$\mu_{\beta,i} = \frac{1}{n_{CPG}}\sum_{j=1}^{500} \beta_{j,i} \mathrm{I}[!\mathrm{is.na}(\beta_{j,i})]$
and
$n_{CPG} = \sum_{j=1}^{500} \mathrm{I}[!\mathrm{is.na}(\beta_{j,i})]$.

We remove tumors with $> 5\%$ missing values
(`r sum(colMeans(is.na(brca.methyl)) > 0.05)` tumors)

```{r remove_missing_data}
too_many_na_values <- colMeans(is.na(brca.methyl)) >= 0.05

brca.methyl <- brca.methyl[, !too_many_na_values]
brca.seq <- brca.seq[, !too_many_na_values]
cibersort <- cibersort[!too_many_na_values, ]
purity <- purity[!too_many_na_values, ]
purity$LUMP.bin <- cut(purity$LUMP, breaks = seq(0.7, 1.0, 0.1), include.lowest = TRUE)
cat(" dimensions of brca.methyl  ", dim(brca.methyl), "\n",
    "dimensions of brca.seq     ", dim(brca.seq), "\n",
    "dimensions of cibersort    ", dim(cibersort), "\n",
    "dimensions of purity       ", dim(purity), "\n")
cat("Names align:", isTRUE(all.equal(getTumorIDs(colnames(brca.methyl)),
                                     getTumorIDs(colnames(brca.seq)))),
    isTRUE(all.equal(getTumorIDs(colnames(brca.methyl)),
                     getTumorIDs(cibersort$Mixture))),
    isTRUE(all.equal(getTumorIDs(colnames(brca.methyl)),
                     getTumorIDs(purity$Sample.ID))), "\n")
```

```{r measured_s_beta}
s_beta_measured <- apply(brca.methyl, 2L, sd, na.rm = TRUE)
c_beta_measured <- 1.0 - s_beta_measured
data_measured <- data.frame("purity" = purity$LUMP,
                            "purity.bin" = purity$LUMP.bin,
                            "c_beta" = c_beta_measured,
                            "model" = rep("measured", length(c_beta_measured)))

data_models <- data_measured
```

#### Correlation with purity

```{r echo = FALSE}
data_measured <- data.frame("purity" = purity$LUMP,
                            "purity.bin" = purity$LUMP.bin,
                            "c_beta" = c_beta_measured,
                            "model" = rep("measured", length(c_beta_measured)))

data_models <- data_measured

gg <- ggplot(data = data_models, aes(c_beta_measured, purity)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm") +
  ggpubr::stat_cor() +
  gg_theme +
  ggtitle(expression("Measured c"[beta] ~ "vs LUMP")) +
  xlab(expression("Measured c"[beta])) + ylab("LUMP")
print(gg)
```

### Model ($c_{{\widehat{\beta}}}^{\alpha})$ {.tabset}

#### Definition

The average beta value at each CpG site is defined as

$$
\beta_{k}^{t} = \frac{1}{p}\left[\beta_{k}^{m} - (1 - p) ~ \beta^{imp}_{k}\right],
$$

where, as before, $p$ is the tumor purity, $\beta_{k}^{m}$ is the
measured beta value at CpG site $k$. In this adjustment, so estimate
$\beta_{k}^{imp}$ as

$$
\beta^{imp}_{k} = \sum_{l = 1}^{22} \frac{s_{l} * \beta_{l,k}}{S}.
$$ Here, $s_{l}$ is the absolute score of the $l^{th}$ cell-type and $S$
is the absolute score of the tumor, both taken from the CIBERSORT
analysis; and $\beta_{l,k}$ is the average methylation for CpG site $k$
in the $l^{th}$ cell-type taken from the WBC lookup table.

#### Beta Value Lookup

```{r read_wbc_xlsx}
lookup <- readxl::read_excel(file.path(save_path, params$lookup_filename))

# keep only the 500 CpG sites of interest
idx <- match(rownames(brca.methyl), lookup$ID_REF)
if (any(is.na(idx))) stop("CpG sites don't match up")

lookup <- lookup[idx, ]
cat("All selected fCpG sites are present", 
    isTRUE(all.equal(lookup$ID_REF, rownames(brca.methyl))), "\n")
```

Manipulate these results into the structure of CIBERSORTx.

```{r convert_table_to_LM22}
B.cells <- c("B.cells.memory", "Plasma.cells", 
             "B.cells.naive")
Macro <- c("Macrophages.M2", "Macrophages.M0", 
           "Dendritic.cells.activated", "Monocytes", 
           "Macrophages.M1", "Dendritic.cells.resting")
Mast <- c("Eosinophils", "Mast.cells.activated", "Mast.cells.resting")
Myeloid <- c("Neutrophils")
NK <- c("NK.cells.resting", "NK.cells.activated")
T.CD8.cells <- c("T.cells.gamma.delta", "T.cells.CD8")
T.CD4.cells <- c("T.cells.follicular.helper", "T.cells.CD4.naive", 
                 "T.cells.regulatory..Tregs.", "T.cells.CD4.memory.activated", 
                 "T.cells.CD4.memory.resting")

grouping_to_wbc <- list("CD4+ T cells" = T.CD4.cells,
                        "CD8+ T cells" = T.CD8.cells,
                        "CD14+ Monocytes" = Macro,
                        "CD19+ B cells" = B.cells,
                        "CD56+ NK cells" = NK,
                        "Eosinophils" = Mast,
                        "Neutrophils" = Myeloid)
cat("All cell types are present in lookup", 
    all(names(grouping_to_wbc) %in% colnames(lookup)), "\n")

idx <- sapply(seq_along(grouping_to_wbc),
              function(i) rep(names(grouping_to_wbc)[i], 
                              length(grouping_to_wbc[[i]]))) |> unlist()
lookup_lm22 <- lookup[, idx]
colnames(lookup_lm22) <- unlist(grouping_to_wbc)
lookup_lm22 <- data.matrix(lookup_lm22)
rownames(lookup_lm22) <- lookup$ID_REF

tst <- NULL
for (i in 1L:length(grouping_to_wbc)) {
  for (j in 1L:length(grouping_to_wbc[[i]])) {
    tst <- c(tst, isTRUE(all.equal(unlist(lookup[, names(grouping_to_wbc)[i]]),
                                   lookup_lm22[, grouping_to_wbc[[i]][j]],
                                   check.attributes = FALSE)))
  }
}
cat("This realignment worked:", all(tst), "\n")
```

#### Calculate site specific beta values

```{r}
idx <- match(colnames(cibersort[, 2L:23L]), colnames(lookup_lm22))
if (any(is.na(idx))) stop("cell types do not match")

# reorder lookup table to match ordering of cibersort results
lookup_lm22 <- lookup_lm22[, idx]
cat("columns of cibersort align with lookup_lm22", 
    isTRUE(all.equal(colnames(cibersort)[2L:23L], colnames(lookup_lm22))), "\n")
cat("tumor names of cibersort align with brca.methyl", 
    isTRUE(all.equal(getTumorIDs(cibersort$Mixture), 
                     getTumorIDs(colnames(brca.methyl)))), "\n")

# calculate beta_i^imp
lm22_wgts <- matrix(0.0, nrow = nrow(brca.methyl), ncol = ncol(brca.methyl))
for (i in 1L:ncol(brca.methyl)) {
  tumor_index <- which(getTumorIDs(cibersort$Mixture) == getTumorIDs(colnames(brca.methyl)[i]))
  cs_tumor <- cibersort[tumor_index, 2L:23L]
  lm22_wgts[, i] <- {lookup_lm22 %*% t(data.matrix(cs_tumor))} / cibersort$Absolute.score..sig.score.[tumor_index]
}
```

```{r}
beta_adj1_tumor <- {{t(brca.methyl) - t(lm22_wgts) * {1.0 - purity$LUMP}} / purity$LUMP} |> t()

cat("This adjustment results in ", mean(beta_adj1_tumor < 0.0 | beta_adj1_tumor > 1.0, na.rm = TRUE)*100,
    "% of the non-NA beta values outside of [0, 1]; we truncate these.\n")
# truncation to physical values
beta_adj1_tumor <- pmin(pmax(beta_adj1_tumor, 0.0), 1.0)

s_beta_adj1 <- apply(beta_adj1_tumor, 2L, sd, na.rm = TRUE)
c_beta_adj1 <- 1.0 - s_beta_adj1

data_adj1 <- data.frame("purity" = purity$LUMP,
                        "purity.bin" = purity$LUMP.bin,
                        "c_beta" = c_beta_adj1,
                        "model" = rep("adj1", length(c_beta_adj1)))
data_models <- rbind(data_models,
                     data_adj1)
```

#### Correlation with purity and previous estimators

The Pearson Correlation coefficient with

```{r}
cor(cbind("measured" = c_beta_measured,
          "purity_adjusted" = c_beta_adj1,
          "Purity" = purity$LUMP))
```

```{r echo = FALSE}
gg <- ggplot(data = data_adj1, aes(purity, c_beta)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm") +
  ggpubr::stat_cor() +
  gg_theme +
  xlab(expression("Purity-adjusted c"[beta]^alpha)) +
  ylab("LUMP")
print(gg)

data_c_beta <- data.frame("purity" = data_measured$purity,
                          "purity.bin" = data_measured$purity.bin,
                          "measured" = data_measured$c_beta,
                          "adj1" = data_adj1$c_beta)
gg <- ggplot(data = data_c_beta, aes(measured, adj1)) +
  geom_point(aes(color = purity.bin)) +
  geom_smooth(formula = y ~ x, method = "lm") +
  ggpubr::stat_cor() +
  gg_theme +
  ylab(expression("Purity-adjusted c"[beta]^alpha)) +
  xlab(expression("Measured c"[beta])) +
  geom_line(aes(measured, measured))
print(gg)
```

The black line is $c_{\beta}$ vs $c_{\beta}$ showing that this
adjustment shifts all $c_{\beta}$ to lower values.

```{r}
summary(c_beta_measured - c_beta_adj1)
```

## Gene Expressions {.tabset}

### Pearson Correlations {.tabset}

#### Immune Specific Genes

Note that genes with all 0's have been removed.

```{r echo = FALSE}
#there are some zero rows
tst <- rowSums(brca.seq) <= 0.0
brca.seq <- brca.seq[!tst, ]
brca.seq_ensembl <- brca.seq_ensembl[!tst]
brca.seq_gene.name <- brca.seq_gene.name[!tst]

genes_of_interest <- c("CD3D", "CD3E", "CD3G", 
                       "CD4", "CD68", "CD8A", 
                       "CD8B", "FOXP3", "MS4A1", 
                       "NCAM1", "PTPRC")
idx_genes <- match(genes_of_interest, brca.seq_gene.name)

# Calculate Pearson Correlation
result_pearson <- apply(brca.seq[idx_genes, ], MARGIN = 1L, FUN = cor, 
                        y = cbind(c_beta_measured, c_beta_adj1)) |> t()
```

```{r}
immune_correlations <- cbind(brca.seq_gene.name[idx_genes], result_pearson)
colnames(immune_correlations) <- c("Gene Name", "Measured", "Adj1")
rownames(immune_correlations) <- NULL
immune_correlations <- as.data.frame(immune_correlations)
print(immune_correlations)
```

### CIBERSORT {.tabset}

```{r}
# add purity information to cibersort result
idx <- match(getTumorIDs(cibersort$Mixture), purity$Sample.ID)
cibersort <- cibersort[idx, ]
cibersort$purity <- purity$LUMP
cibersort$purity.bin <- purity$LUMP.bin

# add c-beta estimates to cibersort result
cat("cibersort tumor names align with c_beta names", 
    all.equal(getTumorIDs(cibersort$Mixture), getTumorIDs(names(c_beta_measured))),
    "\n")
cibersort$c.beta.measured <- c_beta_measured
cibersort$c.beta.adj1 <- c_beta_adj1
```

#### Absolute Score

```{r plot_absolute_score}
absolute_score_figure <- function(data, x, xname) {
  
  label <- parse(text = paste('"', xname))
  gg <- ggplot(data = subset(data, !is.na(data[[x]])),
               aes(.data[[x]], .data$Absolute.score..sig.score.)) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x, color = "blue") +
    ggpubr::stat_cor() +
    gg_theme +
    ylab("Absolute score") + xlab(label)
  
  print(gg)
}
```

```{r Absolute_score_vs_purity}
table(cibersort$purity.bin, useNA = "ifany")
absolute_score_figure(data = cibersort, x = "purity", xname = 'LUMP"')
```

```{r Absolute_score_vs_Measured}
cibersort$c.beta.measured.bin <- cut(cibersort$c.beta.measured, 
                                     breaks = seq(0.65, 0.9, 0.05),
                                     include.lowest = TRUE)
table(cibersort$c.beta.measured.bin, useNA = "ifany")
absolute_score_figure(data = cibersort, x = "c.beta.measured", xname = 'Measured c"[beta]')
```

```{r Absolute_score_vs_Adj1}
cibersort$c.beta.adj1.bin <- cut(cibersort$c.beta.adj1, 
                                 breaks = seq(0.65, 0.85, 0.05),
                                 include.lowest = TRUE)
table(cibersort$c.beta.adj1.bin, useNA = "ifany")
absolute_score_figure(data = cibersort, x = "c.beta.adj1", xname = 'Purity-adjusted c"[beta]^alpha')
```

#### Cell Fractions

```{r}
metrics <- c("purity", "c.beta.measured", "c.beta.adj1")
all_correlations <- apply(cibersort[, metrics],
                          2L, 
                          function(cbeta) {
                            apply(cibersort[, 2L:23L] / cibersort$Absolute.score..sig.score.,
                                  2L, 
                                  function(x, y) {cor.test(x, y)$estimate}, y = cbeta)
                            }) |> as.data.frame()
all_correlations[order(all_correlations$c.beta.measured), ]
```

```{r}
B.cells <- c("B.cells.memory", "Plasma.cells", 
             "B.cells.naive")
Macro <- c("Macrophages.M2", "Macrophages.M0", 
           "Dendritic.cells.activated", "Monocytes", 
           "Macrophages.M1", "Dendritic.cells.resting")
Mast <- c("Mast.cells.activated", "Mast.cells.resting")
Myeloid <- c("Eosinophils", "Neutrophils")
NK <- c("NK.cells.resting", "NK.cells.activated")
T.cells <- c("T.cells.follicular.helper", "T.cells.CD4.naive", 
             "T.cells.regulatory..Tregs.", "T.cells.CD4.memory.activated", 
             "T.cells.gamma.delta", "T.cells.CD4.memory.resting", 
             "T.cells.CD8")
grouping_determined_by_team <- list("B.cells" = B.cells,
                                    "Macro" = Macro,
                                    "Mast" = Mast,
                                    "Myeloid" = Myeloid,
                                    "NK" = NK,
                                    "T.cells" = T.cells)

totals <- NULL
for (i in seq_along(grouping_determined_by_team)) {
  use <- colnames(cibersort) %in% grouping_determined_by_team[[i]]
  totals <- cbind(totals, rowSums(cibersort[, use, drop = FALSE]))
}
colnames(totals) <- names(grouping_determined_by_team)

df <- data.frame("cell.type" = factor(rep(colnames(totals), each = nrow(cibersort))),
                 "fraction" = c(totals / cibersort$Absolute.score..sig.score.),
                 "purity.bin" = rep(cibersort$purity.bin, ncol(totals)),
                 "c.beta.measured.bin" = rep(cibersort$c.beta.measured.bin, ncol(totals)),
                 "c.beta.adj1.bin" = rep(cibersort$c.beta.adj1.bin, ncol(totals)))

table(df$cell.type, useNA = "ifany")
```

```{r plot_cell_fractions}
cell_fraction_figure <- function(data, x, xname) {
  
  col_jco <- pal_jco("default")(7L)

  colors <- c("B.cells" = col_jco[1L], 
              "Macro" = col_jco[2L], 
              "Mast" = col_jco[3L], 
              "Myeloid" = col_jco[4L], 
              "NK" = col_jco[5L], 
              "T.cells" = col_jco[7L])

  title <- parse(text = paste('"Immune Compartment Fractions vs ', xname))
  label <- parse(text = paste('"', xname))
  gg <- ggplot(data = subset(data, !is.na(data[, x])),
               aes(.data[[x]], fraction, fill = cell.type)) +
    geom_boxplot() +
    scale_fill_manual(values = colors) +
    gg_theme +
    ylab("Fraction of immune compartment") + xlab(label)
  print(gg)
}
```

```{r generate_cell_fraction_plots}
cell_fraction_figure(df, "c.beta.measured.bin", 'Measured c"[beta]')
cell_fraction_figure(df, "c.beta.adj1.bin", 'Purity-adjusted c"[beta]^alpha')
```

