---
title: "Purity Adjustment"
author: "Shannon T. Holloway"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
params:
  save_path: "TARGET"
  methyl_filename: "cohort1.methyl"
  processed_clinical_filename: "clinical.processed.tsv"
  lookup_filename: "GSE35069_series_matrix450bloodsubsets_lookuplist_short_header.xlsx"
  cibersort_filename: "Cibersort_LM22_500perm_Analysis_Tumors.csv"
  output_filename: "cohort1.analysis_tumors.methyl_adjusted.tsv"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
library(jsonlite)

source('util.R')

consts <- read_json(file.path(repo_dir, 'config.json'))

if (consts[["Windows"]]) {
  official_indir <- "P:"
} else {
  official_indir <- consts[['official_indir']]
}
save_path <- file.path(official_indir, params$save_path)
```

#  {.tabset}

## Data {.tabset}

### Read TARGET Methylation Data

```{r read_methylation_from_gdc}
methyl_data <- tryCatch(readRDS(file.path(save_path,
                                          paste0(params$methyl_filename, ".rds"))),
                        error = function(e) {
                          stop("must recreate the tumor methylation object")
                          })
cat("Dimensions of raw methylation data", dim(methyl_data), "\n")
```

### Restrict to Analysis Tumors

We keep only the tumors in the analysis cohort that have bulk RNA data, which
are the tumors included in the CIBERSORT analysis.

```{r restrict_to_analysis_tumors}
clinical <- read.table(file.path(save_path,
                                 params$processed_clinical_filename),
                       sep = "\t", header = TRUE)
clinical$in_analysis_dataset <- as.logical(clinical$in_analysis_dataset)
analysis_tumors <- clinical$sampleID[clinical$in_analysis_dataset]

cibersort <- read.csv(file.path(save_path,
                                params$cibersort_filename))

methyl_tumors <- getTumorIDs(colnames(methyl_data))
keep <- methyl_tumors %in% analysis_tumors &
  methyl_tumors %in% getTumorIDs(cibersort$Mixture)
methyl_data <- methyl_data[, keep]

cat("Tumors in the analysis cohort", length(analysis_tumors), "\n")
cat("Dimensions of current methylation data", dim(methyl_data), "\n")
```

### Read CIBERSORTx Results

The CIBERSORT analysis was completed for the tumors in the analysis cohort with
bulk RNA data using the LM22 signature, B-mode batch correction, absolute
mode, quantile normalization disabled, and 500 permutations.

```{r cibersort_read}
idx <- match(getTumorIDs(colnames(methyl_data)), getTumorIDs(cibersort$Mixture))
cibersort <- cibersort[idx[!is.na(idx)], ]
cat("All current tumors are contained in the CIBERSORT results",
    isTRUE(all.equal(getTumorIDs(cibersort$Mixture), getTumorIDs(colnames(methyl_data)))),
    "\n")
cat("Dimensions of CIBERSORT results", dim(cibersort), "\n")
```

### Purity Values

Purity estimates are the LUMP values in the processed clinical table. The
LUMP method is taken from Aran, D., Sirota, M. and Butte, A. J. (2015)
Systematic pan-cancer analysis of tumor purity. Nature Communications,
6, 8971.

```{r read_purity}
# keep only the data for tumors used in the analysis
idx <- match(getTumorIDs(colnames(methyl_data)), clinical$sampleID)
purity <- data.frame("Sample.ID" = clinical$sampleID[idx],
                     "LUMP" = clinical$LUMP[idx])
cat("Dimensions of purity results", dim(purity), "\n")

cat("All current tumors are in the purity dataset:",
    isTRUE(all.equal(purity$Sample.ID, getTumorIDs(colnames(methyl_data)))), "\n")
```

### Read CpG Sites

```{r clock_set_data}
# subset the methylation data to the clock set fCpGs
cpg_sites <- read.csv(file.path(repo_dir, "3. Select fCpGs", "outputs",
                                "NBL_Clock_CpGs.txt"),
                      header = FALSE)[, 1L]

methyl_data <- methyl_data[rownames(methyl_data) %in% cpg_sites, ]
cat("Dimensions of current methylation data", dim(methyl_data), "\n")
```

## Adjust $\beta$ Values {.tabset}

#### Definition

The average beta value at each CpG site is defined as

$$
\beta_{k}^{t} = \frac{1}{p}\left[\beta_{k}^{m} - (1 - p) ~ \beta^{imp}_{k}\right],
$$

where $p$ is the tumor purity, $\beta_{k}^{m}$ is the
measured beta value at CpG site $k$. In this adjustment, we estimate
$\beta_{k}^{imp}$ as

$$
\beta^{imp}_{k} = \sum_{l = 1}^{22} \frac{s_{l} * \beta_{l,k}}{S}.
$$ Here, $s_{l}$ is the absolute score of the $l^{th}$ cell-type and $S$
is the absolute score of the tumor, both taken from the CIBERSORT
analysis; and $\beta_{l,k}$ is the average methylation for CpG site $k$
in the $l^{th}$ cell-type taken from the WBC lookup table.

#### Beta Value Lookup

```{r read_wbc_xlsx}
# the lookup table is in the "Other" directory of official_indir
lookup <- readxl::read_excel(file.path(official_indir, "Other", params$lookup_filename))

# keep only the CpG sites of interest
idx <- match(rownames(methyl_data), lookup$ID_REF)
if (any(is.na(idx))) stop("CpG sites don't match up")

lookup <- lookup[idx, ]
cat("All selected CpG sites are present",
    isTRUE(all.equal(lookup$ID_REF, rownames(methyl_data))), "\n")
```

Manipulate these results into the structure of CIBERSORTx.

```{r convert_table_to_LM22}
B.cells <- c("B.cells.memory", "Plasma.cells",
             "B.cells.naive")
Macro <- c("Macrophages.M2", "Macrophages.M0",
           "Dendritic.cells.activated", "Monocytes",
           "Macrophages.M1", "Dendritic.cells.resting")
Mast <- c("Eosinophils", "Mast.cells.activated", "Mast.cells.resting")
Myeloid <- c("Neutrophils")
NK <- c("NK.cells.resting", "NK.cells.activated")
T.CD8.cells <- c("T.cells.gamma.delta", "T.cells.CD8")
T.CD4.cells <- c("T.cells.follicular.helper", "T.cells.CD4.naive",
                 "T.cells.regulatory..Tregs.", "T.cells.CD4.memory.activated",
                 "T.cells.CD4.memory.resting")

# align these with the headers from the WBC datafile
grouping_to_wbc <- list("CD4+ T cells" = T.CD4.cells,
                        "CD8+ T cells" = T.CD8.cells,
                        "CD14+ Monocytes" = Macro,
                        "CD19+ B cells" = B.cells,
                        "CD56+ NK cells" = NK,
                        "Eosinophils" = Mast,
                        "Neutrophils" = Myeloid)
cat("All cell types are present in lookup",
    all(names(grouping_to_wbc) %in% colnames(lookup)), "\n")
cat("All cell types are present in cibersort",
    all(unlist(grouping_to_wbc) %in% colnames(cibersort)), "\n")

idx <- sapply(seq_along(grouping_to_wbc),
              function(i) rep(names(grouping_to_wbc)[i],
                              length(grouping_to_wbc[[i]]))) |> unlist()
lookup_lm22 <- lookup[, idx]
colnames(lookup_lm22) <- unlist(grouping_to_wbc)
lookup_lm22 <- data.matrix(lookup_lm22)
rownames(lookup_lm22) <- lookup$ID_REF

tst <- NULL
for (i in 1L:length(grouping_to_wbc)) {
  for (j in 1L:length(grouping_to_wbc[[i]])) {
    tst <- c(tst, isTRUE(all.equal(unlist(lookup[, names(grouping_to_wbc)[i]]),
                                   lookup_lm22[, grouping_to_wbc[[i]][j]],
                                   check.attributes = FALSE)))
  }
}
cat("This realignment worked:", all(tst), "\n")
```

#### Calculate site specific beta values

```{r calculate_weight}
# align the cell type columns
idx <- match(colnames(cibersort[, 2L:23L]), colnames(lookup_lm22))
if (any(is.na(idx))) stop("cell types do not match")

# reorder lookup table to match ordering of cibersort results
lookup_lm22 <- lookup_lm22[, idx]
cat("columns of cibersort align with lookup_lm22",
    isTRUE(all.equal(colnames(cibersort)[2L:23L], colnames(lookup_lm22))), "\n")
cat("tumor names of cibersort align with methyl_data",
    isTRUE(all.equal(getTumorIDs(cibersort$Mixture),
                     getTumorIDs(colnames(methyl_data)))), "\n")

# calculate beta_i^imp
lm22_wgts <- matrix(0.0, nrow = nrow(methyl_data), ncol = ncol(methyl_data))
for (i in 1L:ncol(methyl_data)) {
  tumor_index <- which(getTumorIDs(cibersort$Mixture) == getTumorIDs(colnames(methyl_data)[i]))
  cs_tumor <- cibersort[tumor_index, 2L:23L]
  lm22_wgts[, i] <- {lookup_lm22 %*% t(data.matrix(cs_tumor))} / cibersort$Absolute.score..sig.score.[tumor_index]
}
```

```{r generate_and_save}
beta_adj1_tumor <- {{t(methyl_data) - t(lm22_wgts) * {1.0 - purity$LUMP}} / purity$LUMP} |> t()

cat("This adjustment results in ", mean(beta_adj1_tumor < 0.0 | beta_adj1_tumor > 1.0, na.rm = TRUE)*100,
    "% of the non-NA beta values outside of [0, 1]; we truncate these.\n")
# truncation to physical values
beta_adj1_tumor <- pmin(pmax(beta_adj1_tumor, 0.0), 1.0)

write.table(beta_adj1_tumor,
            file = file.path(save_path, params$output_filename),
            sep = "\t")
```
